Lossy compression of DNA sequencing quality data

### 2015-05-14 manuscript/q-compression.tex
32 bytes => 64 bytes.
6 coefficients => 8 coefficients.

If quality values are modeled, compression is conceivably possible by
replacing the original quality values by a representation of the
model. For example, quality values can be conceptualized as bivariate
data with the ordered nucleotides (1 to read length) representing the
abscissa, and quality values representing the ordinate. We model read
quality values as polynomial functions obtained with least-squares
fitting, as one approach to the compression of read quality values by
modeling. Although polynomial functions have significantly fewer
parameters than a read-length string of raw quality values (e.g., one
to six coefficients in our approach), the use of floating point
numbers to store coefficients greatly limits the compression potential
of the models. The coefficients could be represented with fewer bits,
perhaps with some loss of precision, but we have not considered
alternative representations in this study.

@@

As the degree of the polynomial increases, the bits/base-call increase
and the \textsc{mse} decreases at an exponential rate. The 7th-degree
polynomial regression results in the largest bits/base-call, as it
requires 32 bytes per read before subsequent compression with
\textsc{bz}ip2. For the \textit{Mus musculus} data set, which has read
lengths of only 36 nt, the memory required for storage, even after
subsequent compression with \textsc{bz}ip2 exceeds that for the
original \textsc{ascii} quality values.

>>>>

If quality values are modeled, compression is conceivably possible by
replacing the original quality values by a representation of the
model. For example, quality values can be conceptualized as bivariate
data with the ordered nucleotides (1 to read length) representing the
abscissa, and quality values representing the ordinate. We model read
quality values as polynomial functions obtained with least-squares
fitting, as one approach to the compression of read quality values by
modeling. Although polynomial functions have significantly fewer
parameters than a read-length string of raw quality values (e.g., one
to eight coefficients in our approach), the use of double-precision floating-point
numbers to store coefficients greatly limits the compression potential
of the models. The coefficients could be represented with fewer bits,
perhaps with some loss of precision, but we have not considered
alternative representations in this study.

@@

As the degree of the polynomial increases, the bits/base-call increase
and the \textsc{mse} decreases at an exponential rate. The 7th-degree
polynomial regression results in the largest bits/base-call, as it
requires 64 bytes per read before subsequent compression with
\textsc{bz}ip2. For the \textit{Mus musculus} data set, which has read
lengths of only 36 nt, the memory required for storage, even after
subsequent compression with \textsc{bz}ip2 exceeds that for the
original \textsc{ascii} quality values.
